\bf Distortion corrected kernel density estimator on Riemannian manifolds

### 2024-09-11 jcgs/03-chap3-kderm.tex
minor changes

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
%\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%% BIBLIOGRAPHY

\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}
\ExecuteBibliographyOptions{bibencoding=utf8, minnames=1, maxnames=3, maxbibnames=99,dashed=false, terseinits=true, giveninits=true, uniquename=false, uniquelist=false,doi=false, isbn=false, url=true, sortcites=false, date=year}

\addbibresource{kderm.bib}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{url}{}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here

% Pandoc citation processing

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{tabu,algorithmic}
\usepackage[section]{placeins}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\urlstyle{tt}  % use monospace font for urls
\usepackage[cal = txupr]{mathalpha}
\usepackage{geometry}
\geometry{margin=2.2cm}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage{amsfonts,amssymb,bm}
\usepackage{mathpazo}
\usepackage[no-weekday]{eukdate}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\DeclareMathOperator\supp{supp}

\allowdisplaybreaks
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.9}

\usepackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\def\spacingset#1{\renewcommand{\baselinestretch}%
  {#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
  {
    \title{\bf Distortion corrected kernel density estimator on Riemannian manifolds}

    \author{
            Fan Cheng \\
      Monash University\\
       and \\       Rob J Hyndman \thanks{Rob J Hyndman's research is partially funded by the Australian Government through the Australian Research Council Industrial Transformation Training Centre in Optimisation Technologies, Integrated Methodologies, and Applications (OPTIMA), Project ID IC200100009.} \\
      Monash University\\
       and \\       Anastasios Panagiotelis \thanks{Corresponding Author} \\
      The University of Sydney\\
          }
    \maketitle
  } \fi

\if1\blind
  {
    \bigskip
    \bigskip
    \bigskip
    \begin{center}
      {\LARGE\bf Distortion corrected kernel density estimator on Riemannian manifolds}
    \end{center}
    \medskip
  } \fi

\bigskip
\begin{abstract}
  Manifold learning obtains a low-dimensional representation of an underlying Riemannian manifold supporting high-dimensional data. Kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding, with a bandwidth that depends on the estimated Riemannian metric at each data point. Exploiting the geometric information of the manifold leads to more accurate density estimation, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations including one with data lying in a 100 dimensional ambient space. We demonstrate that the proposed DC-KDE improves the density estimates as long as the manifold learning embedding is of sufficient quality, and has higher rank correlations with the true manifold density. Further simulation results are provided via a supplementary R shiny app. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data.
\end{abstract}

\noindent%
{\it Keywords:} manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
  \vfill

  \newpage
  \spacingset{1.45} % DON'T change the spacing!

@@

Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Scott2015-vl} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}. For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\). The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da}, and \textcite{Berenfeld2022-fd}. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\). Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx}, and \textcite{Cheng2021-dh}. One could easily apply KDE to the output vectors \(\bm{y}\). However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
\[Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,\]
where \(f\) is the probability density function on the manifold, \(\bm{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a change of variables, this would imply the following density for \(\bm{y}\),
\begin{equation}
\label{eq:changevar}
Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\bm{H}(\bm{y})|^{1/2}d\bm{y},
\end{equation}
where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\). It can be seen that the density in terms of \(\bm{y}\), differs from the density on the manifold due to the term \(|\bm{H}(\bm{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant. Since \(\psi\) will typically be a non-linear function, the value of \(|\bm{H}(\bm{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\bm{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\bm{y}\) is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data. Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm. \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space. \autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage. Conclusions and discussions are presented in \autoref{dckdeconclusion}. Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.

@@

In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.

In the following, we denote \(M\) as the \(d\)-dimensional manifold from which our data are sampled. Points on this manifold are denoted \(\bm{p}\) in general, with \(\bm{p}_1,\dots,\bm{p}_n\) denoting the observed sample. Often \(\bm{p}_i\) will be high-dimensional vectors such that \(\bm{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this need not be the case. For instance, \(\bm{p}_i\) may be probability distributions on a statistical manifold. The methods we propose for estimating the density at each \(\bm{p}_i\) only require some sense of distance between the `input' points, \(d(\bm{p}_i,\bm{p}_j)\), such that we can apply dimension reduction algorithms to obtain an `output' embedding, \(\bm{y}_1,\dots,\bm{y}_n\), where \(\bm{y}_i\in\mathbb{R}^d\). We will denote this embedding as \(\bm{y}_i=\psi(\bm{p}_i)\). Finally, we denote by \(\lambda\) the Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual Euclidean norm and following \textcite{Pelletier2005-vu}, we make the following assumptions about the kernel function \(K:\mathbb{R}_+\rightarrow\mathbb{R}\),
\begin{equation}
\begin{aligned}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=1;
(ii) \int_{\mathbb{R}^{d}} \bm{y} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\bm{y}\|^{2} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\bm{y}\|)=K(0).
\end{aligned}
\end{equation}

@@

The intuition behind this estimator is very clear for a uniform kernel. The density at a point \(\bm{y}\) is equal to the proportion of sample points that lie within a ball of radius \(r\) centered at \(\bm{y}\), times a term that ensures the density integrates to 1. In general, the bandwidth matrix need not be proportional to the identity matrix. However, the intuition remains the same, only that the ball of radius \(r\) centered at \(\bm{y}\) is found with respect to Mahalanobis distance rather than the usual Euclidean distance. For more on kernel density estimation in the Euclidean case, see \textcite{Scott2015-vl} and references therein. Kernel density estimators of this form can and have been applied directly on the output embedding \(\bm{y}\), and we will consider this approach as a benchmark in \autoref{simulation}.

@@

For kernel density estimation on a known manifold, \textcite{Pelletier2005-vu} propose the following estimator,
\begin{equation}
\label{eq:denriem}
\hat{f}(\bm{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\bm{p}_i}(\bm{p})} K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right),
\end{equation}

where \(d_g(\bm{p}, \bm{p}_i)\) denotes the geodesic distance between two points on the manifold \(\bm{p}\) and \(\bm{p}_i\) and \(\theta_{\bm{p}_i}(\bm{p})\) is known as the volume density function. The intuition behind the term \(K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)\) is relatively clear. For example, for a uniform kernel, the estimator at point \(\bm{p}\) will still depend on the proportion of sample points within a ball of radius \(r\) centered at \(\bm{p}\). However, in this case, the geodesic distance on the manifold is used, rather than Euclidean or Mahalanobis distance. An additional technical assumption is that \(r\) is less than the injectivity radius of the manifold. A definition of the injectivity radius is given by \textcite{Chavel2006-mp} and also provided in the appendix. For our purposes, it is sufficient to note that this assumption precludes the possibility that the radius of a ball around \(\bm{p}\) is so large that some points `fall inside' the ball more than once. For example on a sphere, a ball with a radius greater than half the circumference of a great circle will wrap back around the sphere. This phenomenon also explains why the kernel function must be bounded for density estimation on manifolds.

@@

More formally, the volume density function can be explained as follows. Consider the exponential map around \(\bm{p}\), given by \(exp_{\bm{p}}(\bm{q})\), mapping vectors in the tangent space, \(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\). Loosely, \(\bm{v}\) `points' in the direction of the geodesic between \(\bm{p}\) and \(\bm{q}\) and traveling along this geodesic at uniform speed \(\|\bm{v}\|\) takes place in one unit of time. Now, consider a chart \(\varphi\) mapping points in the neighborhood of \(\bm{p}\), via the inverse of the exponential map, to these \(\bm{v}\) vectors, expressed in some local coordinate system. The volume density function is then the square root of the determinant of the Riemannian metric expressed in this coordinate system. In proving that the estimator integrates to 1, the volume density function cancels with the term \(|\bm{H}(\bm{y})|^{1/2}\) in Equation \eqref{eq:changevar}, where \(\psi\) is the logarithmic map around \(\bm{p}_i\). For more on the volume density function, see \textcite{Gallot2004-rc} and \textcite{Le_Brigant2019-lj}.

@@

\caption{The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.}\label{fig:vkde}
\end{figure}

@@

@@

\caption{Scatterplot of the 3-d twin peaks data with the same colors indicating the true density as in \autoref{fig:metadensity}.}\label{fig:twinpeaks3d}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

@@

@@

Multivariate kernel density estimation (KDE see @Parzen1962-gt;
@Scott2015-vl and references therein) is a fundamental tool for
exploratory data analysis and unsupervised learning. Applications of KDE
include finding hot spots of traffic networks in the GIS environment
{[}@Xie2008-eb; @Okabe2009-nb{]}, automatic detection in visual
surveillance systems {[}@Elgammal2002-cw{]}, wind power density
detection {[}@Jeon2012-ac{]} and prime prediction via Twitter messages
{[}@Gerber2014-tq{]}. For \(m\)-dimensional data where \(m\) is even of
moderate size, KDE is challenging due to the well-known curse of
dimensionality. However, the manifold hypothesis suggests that many
high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\)
where \(d\ll m\). The problem of KDE on manifolds has been well studied
in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da,
and @Berenfeld2022-fd. However, in these settings, it is assumed that
the underlying manifold is known, which is rarely true in practice. The
is only a small literature dealing with the case where the manifold is
unknown. Recently, @berenfeld2021-den, @berenfeld2022-est and
@divol2022measure demonstrated the minimax optimality of certain types
of KDE on manifolds. In contrast, our objective in this paper is to
propose a new KDE for data that lie on some unknown manifold found using
embedding techniques.

When dealing with data that lie on a manifold, it is common to use
dimension-reduction techniques. These algorithms, which include ISOMAP,
LLE, t-SNE, and UMAP among others, can be thought of as a mapping
\(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold
\(\bm{p}\) to \(d\)-dimensional vectors \(\bm{y}\), where the dimension
\(d\) is either assumed to be known a priori or needs to be estimated
separately. Details of these algorithms are summarized in
@Cayton2005-dp, @Lee2007-wq, @Izenman2012-mx, and @Cheng2021-dh. One
could easily apply KDE to the output vectors \(\bm{y}\). However, this
would be a misleading representation of the density of points as they
lie on the manifold. To see how, consider that the density on the
manifold must be defined with respect to an appropriate volume form,
that is \[Pr(\bm{p}\in\mathcal{A})=\int_\mathcal{A} f(\bm{p}) dvol_M,\]
where \(f\) is the probability density function on the manifold,
\(\bm{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on
the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a
change of variables, this would imply the following density for
\(\bm{y}\), \begin{equation}
\label{eq:changevar}
Pr(\bm{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\bm{y})|\det \bm{H}(\bm{y})|^{1/2}d\bm{y},
\end{equation} where \(\psi(\mathcal{A})\) is the image of
\(\mathcal{A}\) under \(\psi\) and \(\bm{H}(\bm{y})\) is the Riemannian
metric expressed in local coordinates given by the mapping \(\psi\)
{[}@doCarmo1992-rm{]}. It can be seen that the density in terms of
\(\bm{y}\), differs from the density on the manifold due to the term
\(|\bm{H}(\bm{y})|^{1/2}\) which can be considered roughly analogous to
the Jacobian determinant. Since \(\psi\) will typically be a non-linear
function, the value of \(|\bm{H}(\bm{y})|^{1/2}\) will change across the
manifold, in a way that measures the distortion induced by the dimension
reduction algorithm. Notably, different choices of dimension reduction
algorithm, imply different \(\psi\), meaning that if KDE is applied to
the output \(\bm{y}\) coming from two different dimension reduction
algorithms, this could lead to two very different sets of conclusions
about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the
output of dimension reduction algorithms but accounts for the distortion
induced by the algorithm. For this reason, we refer to it as the
distortion-corrected KDE (DC-KDE). Critical to our estimator, is
obtaining an estimate of \(|\bm{H}(\bm{y})|^{1/2}\) for this purpose we
use the Learn Metric algorithm of @Perrault-Joncas2013-pq which augments
any dimension reduction algorithm with an estimate of the Riemannian
metric at each data point. One particular downstream task that we are
interested in is anomaly detection, where anomalies are identified as
points of low density. In a number of simulation studies, we demonstrate
two benefits that arise when using our new estimator. The first is that
using distortion correction leads to more accurate density estimates and
better detection of anomalies compared to the case where distortion
correction is not used. The second is that when the estimates are found
using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the
results regarding the density estimates and the detection of anomalies
are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or
three-dimensional scatterplot of \(\bm{y}\) to be augmented by coloring
points according to the magnitude of the density estimate, where the
dimension of \(\bm{y}\), \(d\), is set as 2 or 3 in the dimension
reduction algorithms. In this way, the user is able to tell whether a
region that has a cluster of points in the output space \(\bm{y}\) is
truly a high-density region or if instead, this is an artifact of the
dimension reduction algorithm `squeezing' points together. This is
particularly pertinent in the case of t-SNE which can form clusters of
points even when no such clusters exist in the underlying data.
Furthermore, our estimator can be used for manifolds with elements that
are a variety of objects, for instance, rotational frames or images. In
our application we consider a probabilistic manifold whose elements are
densities, meaning that we estimate a density of densities. Finally, our
proposed estimator also takes the form of a KDE with variable bandwidth
meaning that it also contributes to extensive literature on this topic
{[}see Section 6.6 of @Scott2015-vl for a summary{]}.

The rest of the paper is organized as follows. In \autoref{kderm}, we
present our distortion-corrected kernel density estimator for Riemannian
manifolds. Our estimator combines insights from two existing papers, the
first is @Pelletier2005-vu who propose a KDE for data on manifolds that
are known, the second is @Perrault-Joncas2013-pq who propose the Learn
Metric algorithm which we use to estimate the distortion induced by a
dimension reduction algorithm. \autoref{simulation} is composed of two
simulations; the first deals with 2-dimensional data from a Gaussian
mixture model mapped onto a 3-dimensional `twin peaks' manifold and the
second with a 5-dimensional semi-hypersphere data embedded into a
100-dimensional space. \autoref{dckdeapplication} contains an
application to an Irish smart meter dataset that visualizes and
identifies households with an anomalous distribution of energy usage.
Conclusions and discussions are presented in \autoref{dckdeconclusion}.
Some additional background on important terms in Riemannian geometry
required to understand our proposed method is provided in
\autoref{riemgeo}.

@@

In this section, we introduce our method for kernel density estimation
on manifolds that uses an embedding from a dimension reduction algorithm
while correcting for the distortion induced by this embedding. Since
some readers may be unfamiliar with the nuances of manifolds, we first
discuss kernel density estimation for data in Euclidean space, then
illustrate in \autoref{Pellet} how this generalizes to the estimator of
@Pelletier2005-vu, which can be applied when the data lie on some known
manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm
of @Perrault-Joncas2013-pq, which augments an embedding derived from a
dimension reduction algorithm with an estimate of the Riemannian metric
expressed in local coordinates. By combining elements from the work of
@Pelletier2005-vu and @Perrault-Joncas2013-pq, we derive our own novel
distortion-corrected kernel density estimate in \autoref{DCKDE}. To keep
this section as succinct as possible, we do not define concepts such as
manifolds, charts, geodesic distance, etc., but provide this information
for readers unfamiliar with differential geometry in \autoref{riemgeo}.

In the following, we denote \(M\) as the \(d\)-dimensional manifold from
which our data are sampled. Points on this manifold are denoted
\(\bm{p}\) in general, with \(\bm{p}_1,\dots,\bm{p}_n\) denoting the
observed sample. Often \(\bm{p}_i\) will be high-dimensional vectors
such that \(\bm{p}_i\in\mathbb{R}^m\) with \(m\gg d\), however, this
need not be the case. For instance, \(\bm{p}_i\) may be probability
distributions on a statistical manifold. The methods we propose for
estimating the density at each \(\bm{p}_i\) only require some sense of
distance between the `input' points, \(d(\bm{p}_i,\bm{p}_j)\), such that
we can apply dimension reduction algorithms to obtain an `output'
embedding, \(\bm{y}_1,\dots,\bm{y}_n\), where
\(\bm{y}_i\in\mathbb{R}^d\). We will denote this embedding as
\(\bm{y}_i=\psi(\bm{p}_i)\). Finally, we denote by \(\lambda\) the
Lebesgue measure of \(\mathbb{R}^d\), letting \(\|\cdot\|\) be the usual
Euclidean norm and following @Pelletier2005-vu, we make the following
assumptions about the kernel function
\(K:\mathbb{R}_+\rightarrow\mathbb{R}\), \begin{equation}
\begin{aligned}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=1;
(ii) \int_{\mathbb{R}^{d}} \bm{y} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})=0;
(iii) \int_{\mathbb{R}^{d}}\|\bm{y}\|^{2} K(\|\bm{y}\|) \mathrm{d} \lambda(\bm{y})<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(\|\bm{y}\|)=K(0).
\end{aligned}
\end{equation}

@@

The intuition behind this estimator is very clear for a uniform kernel.
The density at a point \(\bm{y}\) is equal to the proportion of sample
points that lie within a ball of radius \(r\) centered at \(\bm{y}\),
times a term that ensures the density integrates to 1. In general, the
bandwidth matrix need not be proportional to the identity matrix.
However, the intuition remains the same, only that the ball of radius
\(r\) centered at \(\bm{y}\) is found with respect to Mahalanobis
distance rather than the usual Euclidean distance. For more on kernel
density estimation in the Euclidean case, see @Scott2015-vl and
references therein. Kernel density estimators of this form can and have
been applied directly on the output embedding \(\bm{y}\), and we will
consider this approach as a benchmark in \autoref{simulation}.

@@

For kernel density estimation on a known manifold, @Pelletier2005-vu
propose the following estimator, \begin{equation}
\label{eq:denriem}
\hat{f}(\bm{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\bm{p}_i}(\bm{p})} K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right),
\end{equation}

where \(d_g(\bm{p}, \bm{p}_i)\) denotes the geodesic distance between
two points on the manifold \(\bm{p}\) and \(\bm{p}_i\) and
\(\theta_{\bm{p}_i}(\bm{p})\) is known as the volume density function.
The intuition behind the term
\(K\left(\frac{d_g(\bm{p}, \bm{p}_i)}{r}\right)\) is relatively clear.
For example, for a uniform kernel, the estimator at point \(\bm{p}\)
will still depend on the proportion of sample points within a ball of
radius \(r\) centered at \(\bm{p}\). However, in this case, the geodesic
distance on the manifold is used, rather than Euclidean or Mahalanobis
distance. An additional technical assumption is that \(r\) is less than
the injectivity radius of the manifold. A definition of the injectivity
radius is given by @Chavel2006-mp and also provided in the appendix. For
our purposes, it is sufficient to note that this assumption precludes
the possibility that the radius of a ball around \(\bm{p}\) is so large
that some points `fall inside' the ball more than once. For example on a
sphere, a ball with a radius greater than half the circumference of a
great circle will wrap back around the sphere. This phenomenon also
explains why the kernel function must be bounded for density estimation
on manifolds.

@@

More formally, the volume density function can be explained as follows.
Consider the exponential map around \(\bm{p}\), given by
\(exp_{\bm{p}}(\bm{v})\), mapping vectors in the tangent space,
\(\bm{v}\in T_{\bm{p}}M\), to points on the manifold, \(\bm{q}\in M\).
Loosely, \(\bm{v}\) `points' in the direction of the geodesic between
\(\bm{p}\) and \(\bm{q}\) and traveling along this geodesic at uniform
speed \(\|\bm{v}\|\) takes place in one unit of time. Now, consider a
chart \(\varphi\) mapping points in the neighborhood of \(\bm{p}\), via
the inverse of the exponential map, to these \(\bm{v}\) vectors,
expressed in some local coordinate system. The volume density function
is then the square root of the determinant of the Riemannian metric
expressed in this coordinate system. In proving that the estimator
integrates to 1, the volume density function cancels with the term
\(|\bm{H}(\bm{y})|^{1/2}\) in Equation @ref(eq:changevar), where
\(\psi\) is the logarithmic map around \(\bm{p}_i\). For more on the
volume density function, see @Gallot2004-rc and @Le\_Brigant2019-lj.

@@

\caption{(ref:vkdecaption)}\label{fig:vkde}
\end{figure}

@@

(ref:twinpeaks3dcaption) Scatterplot of the 3-d twin peaks data with the
same colors indicating the true density as in \autoref{fig:metadensity}.

@@

\caption{(ref:twinpeaks3dcaption)}\label{fig:twinpeaks3d}
\end{figure}
